Is anchoring better?

### 2022-05-25 manuscript/body.tex
Some refactoring and some MS comments (and tiny changes)

@@

In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility. This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest. Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \citep{fox2009effect, leeper2017package, lenth2018package}. For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.

@@

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}. This generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''. More specifically, we compute the expected outcome by meaningfully holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor. On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The grey points are the binned simulated $y$. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three mthods and the ``true'' predictions based on the simulation parameters. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches closely matches the ``true'' predictions. B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A, all the three packages averages non-focal predictors $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular \pkg{emmeans} and \pkg{effects} uses \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\textwidth]{justify_preds.isolate.pdf}
\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, corresponds to the effects of the focal predictor from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center. The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal predictor.}
\label{fig:justify_ci_plots}
\end{figure}

@@

When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions. An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

@@

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{cubic_predictors_preds.ggp.pdf}
\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial. B: The focal predictor is not a polynomial function, implying that the complex interaction is in non-focal predictor. In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but crosses at the mean of the focal predictor (model center) only in the case of simple (non-polynomial) focal predictor -- B. In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A. The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively, while the outer blue curves are the non-anchored confidence bands in both cases. The central curve and trend line, in A and B, respectively, are the predicted household size based on age and wealth index, respectively.}
\label{fig:pred_cubic_plots}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{glm_two_predictor_preds.ggp.pdf}
\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{mediate_preds.ggp.pdf}
\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models. A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, for both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

>>>>

\jd{How did all the figures wind up in back?}

@@

\jd{Can we stake out early what we mean by predictions vs.~effects? It still seems muddled together. I guess it's complicated because in many cases the estimate is the same (only CIs differ). But I think we should do it anyway.}
In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility. This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest. Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \citep{fox2009effect, leeper2017package, lenth2018package}. 
\jd{Are we still saying these (prediction and effects) are the same? Or are we now saying they're different, with the same citations?} For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.

@@

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) \jd{Don't understand what the bins are doing here (when we are holding the reference point constant); it's not what we do, is it?} if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}. This generates \jd{Are predictions and predictor effects really used for the same thing?} -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''. More specifically, we compute the expected outcome by holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor. On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor. 
\jd{I don't understand this: certainly both predictions and effects have uncertainty.}

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point. 
\jd{This is ambiguous; we need to be talking about input variables and model variables by now. Also: what do the packages do about categorical non-focal predictors? Do we know?}
However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
\jd{What are true predictions? Why does it look like effects matches varpred, but it sounds like we're saying it matches emmeans?}

\begin{figure}[!h]
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The grey points are the binned simulated $y$. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three mthods and the ``true'' predictions based on the simulation parameters. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches closely matches the ``true'' predictions. B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

\jd{Why do we say ``similar"? Are they not the same?}
In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A, all the three packages averages non-focal predictors $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular \pkg{emmeans} and \pkg{effects} uses \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, corresponds to the effects of the focal predictor from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center. The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal predictor.}
\label{fig:justify_ci_plots}
\end{figure}

@@

\jd{Hard to parse. Why ``for example''? The key here is that the lm case works out well because it's all about linear averaging (the mean prediction is the prediction from the mean point), whereas the glm case is non-linear averaging. There should be a link also to why we use the ``model center'' (we're using the correct center point for the linear averaging to work out.}
When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions. An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

@@

\jd{Is this borrowed from me? I'm not really following the structure of the MS right now. Maybe I'll be clearer in the morning.}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics{cubic_predictors_preds.ggp.pdf}
\end{center}
\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial. B: The focal predictor is not a polynomial function, implying that the complex interaction is in non-focal predictor. In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but crosses at the mean of the focal predictor (model center) only in the case of simple (non-polynomial) focal predictor -- B. In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A. The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively, while the outer blue curves are the non-anchored confidence bands in both cases. The central curve and trend line, in A and B, respectively, are the predicted household size based on age and wealth index, respectively.}
\label{fig:pred_cubic_plots}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics{glm_two_predictor_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models. A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, for both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}
